Over-relaxed Sinkhorn--Knopp Algorithm for Regularized Optimal Transport

### 2017-10-19 paper.tex
Algo
Details

The computation of optimal transport between two data relies on the estimation of an optimal transport matrix, the entries of which represent the quantity of mass transported between  data locations. 
Regularization of optimal transport with strictly convex regularizers \cite{cuturi13, dessein2016}  nevertheless involves a spreading of the mass. Hence, for particular purposes such as color interpolation \cite{Rabin2014} or gradient flow \cite{2016arXiv160705816C}, it is  necessary  to consider very small regularization of the problem.
In this setting,  the regularized transport problem can be ill-conditioned. This is the issue  we want to tackle here.
Before going into further details, we now briefly introduce the main notations and concepts used all along this article.

@@

In \cite{cuturi13}, it has been proposed to regularize this problem by adding a strictly convex entropy regularization:
\begin{equation}\label{ROT}
\min_{\gamma\in\Ccal_1\cap \Ccal_2\cap \IR^{n_1 n_2}_{+*}}K^\epsilon(\gamma) := \scal{c}{\gamma} 
+ \epsilon KL(\gamma,\mathbf{1})
,\end{equation}
with $\epsilon>0$, $\mathbf{1}$ is the matrix of size $n_1\times n_2$ full of ones and the Kullback-Leibler divergence is:
\begin{equation}\label{KL}
KL(\gamma,\xi) = \sum_{i,j} \gamma_{i,j} \left( \log \left( \frac{\gamma_{i,j}}{\xi_{i,j}} \right) -1  \right) + \sum_{i,j} \xi_{i,j}.
\end{equation}
It was shown in \cite{benamou15}  that the regularized optimal transport matrix $\gamma^*$, which is the unique minimizer of problem \eqref{ROT},  is the Bregman projection of $\gamma^0 = e^{-c/\epsilon}$ onto $\Ccal_1 \cap \Ccal_2$:
\begin{equation}\label{eq:reg_ot_pb}
\gamma^* = \argmin_{\Ccal_1 \cap \Ccal_2} K^\epsilon(\gamma)= P_{\Ccal_1 \cap \Ccal_2} (e^{-c/\epsilon}),
\end{equation}
where $P_{\Ccal}$ is the  Bregman projection onto $\Ccal$ defined as:
\[
P_{\Ccal}(\xi) := \argmin_{\gamma \in \Ccal} KL(\gamma,\xi).
\]

@@

In the discrete setting,  projections actually correspond to diagonal scalings.
\begin{align}\label{scaling}
P_{\Ccal_1}(\gamma) &= \diag(a) \gamma &
a &=  {\mu^1}\oslash{A_1 \gamma} \\
P_{\Ccal_2}(\gamma) &= \gamma \diag(b) &
b &= {\mu^2}\oslash{A_2 \gamma}\nonumber
\end{align}
where $\oslash$ is the pointwise division. 
To compute numerically the solution  one simply has to store $(a^l, b^l)\in\IR^{n_1+n_2}$ and to iterate
\begin{align*}
a^{l+1} &= {\mu^1}\oslash{\gamma^0 b^l} &
b^{l+1} &= {\mu^2}\oslash{^t \gamma^0 a^{l+1}} 
\end{align*}
where the transport matrix can be recovered as :
$\gamma^l = \diag(a^l) \gamma^0 \diag(b^l).$ 
Efficient parallel computations can be considered \cite{cuturi13} and one can even reach real-time computation for large scale problem for certain class of cost matrices $c$ \cite{Solomon2015}. 
For small values of the parameter $\epsilon$, numerical issues can arise and a stabilization of the algorithm is necessary \cite{2016arXiv160705816C}.
The convergence of the process can nevertheless be very slow  in the setting $\epsilon\to 0$.
\subsection{Related works }
The introduction of relaxation variables through heavy ball approaches \cite{POLYAK19641} has recently gained in popularity  to speed up the convergence of algorithms optimizing convex \cite{2014arXiv1412.7457G} or non convex \cite{Zavriev1993,2016arXiv160609070O} problems. Such schemes have also been empirically considered to accelerate the SK algorithm  in \cite{peyre2016quantum,2017arXiv170801955S}. The convergence of these algorithms is nevertheless not studied yet in the context of regularized optimal transport.

@@

As illustrated in Figure \ref{alternate_projections} (a-b), the SK algorithm, that  performs alternate Bregman projections onto the affine sets $\Ccal_1$ and $\Ccal_2$, can be very slow when $\epsilon\to 0$. The idea developped in this paper is to perform over-relaxed projections in order to accelerate the process, as displayed in Figure \ref{alternate_projections} (c).

@@

Let us define the $\omega$-over-relaxed projection operator by:
\begin{equation}\label{eq:def_or_proj}
\log P^\omega_{\Ccal_k}(\gamma) = (1-\omega) \log \gamma + \omega \log P_{\Ccal_k}(\gamma),
\end{equation}
where the logarithm is taken coordinate-wise.
Note that $P_{\Ccal_k}^0$ is the identity, $P_{\Ccal_k}^1 = P_{\Ccal_k}$ is the standard Bregman projection, and $P_{\Ccal_k}^2$ is an involution (this uses the fact that $\Ccal_k$ is an affine subspace).

@@

We argue that it makes sense to consider this Lyapunov function.
Obviously, it is closely related to the notion of Bregman projection that is used throughout the algorithm.
But it is also easy to calculate its decrease, as shown in lemma \ref{lemma:lyapunov_decrease}.
Moreover, in practice it allows the use of a wide range of over-relaxation parameters.

\begin{lemma} \label{lemma:KL_compact}
	For any $M \in \IR_+^*$, the sublevel set $K := \left\{ \gamma \mid F(\gamma) \le M \right\}$ is compact.
\end{lemma}
\begin{proof}
	First note that K is closed in $\IR_{+*}^{n_1 n_2}$, by continuity of $F$.
	For $\gamma \in K$, one has:
	\begin{align*}
	F(\gamma) &= \scal{\gamma^*}{\log \gamma^* - 1} + \scal{\gamma}{1} - \scal{\gamma^*}{\log \gamma}\\
	\sum_{i,j} \gamma_{i,j} - \gamma^*_{i,j} \log \gamma_{i,j} &\le M - \scal{\gamma^*}{\log \gamma^* - 1} .
	\end{align*}
	By concavity, the logarithm admits a simple upper bound: $\log \gamma_{i,j} \le \frac{\gamma_{i,j}}{e}$. Moreover, since $\gamma^*$ is a solution of the problem, each of its coordinates is at most 1.
	\begin{align*}
	\sum_{i,j} \gamma_{i,j} &\le \frac{M - \scal{\gamma^*}{\log \gamma^*-1}}{1-\frac{1}{e}} =: M'.
	\end{align*}
	Therefore each coordinate $\gamma_{i,j}$ is bounded above by $M'$.

	Let us show that the coordinates are bounded below by a strictly positive number; from which it can be deduced that $K$ is bounded and closed in $\IR^{n_1 n_2}$, and therefore compact.
	\begin{align*}
	F(\gamma)
	&\ge \scal{\gamma^*}{\log \gamma^*} - \sum_{i,j} \gamma^*_{i,j} \log(\gamma_{i,j}) - 1\\
	&\ge \scal{\gamma^*}{\log \gamma^*} - \gamma^*_{i_0,j_0} \log(\gamma_{i_0,j_0}) - 1 - \log M',
	\end{align*}
	where the last equation derives from $\log(y_{i,j}) \le \log M'$, and $\sum_{(i,j)\neq(i_0,j_0)} \gamma^*_{i,j} \le 1$.
	Finally:
	\[
	\log(\gamma_{i_0,j_0}) \ge \frac{\scal{\gamma^*}{\log \gamma^*} - M - \log M' - 1}{\gamma^*_{i_0,j_0}},
	\]
	and $K$ is thus compact.
\end{proof}

Note that the difference $F(\gamma) - F(P^\omega_{\Ccal_k}(\gamma))$ may be calculated without knowing $\gamma^*$, as shown by the following lemma.
\begin{lemma}\label{lemma:lyapunov_decrease}
	Take $\gamma$ in $\IR^{mn}_{+*}$. The decrease in value of the Lyapunov function can be calculated with the following formula:
	\begin{equation} \label{eq:kl_diff_scal}
	F(\gamma) - F(P^\omega_{\Ccal_k}(\gamma)) = 
	\scal{\mu^k}{\varphi_\omega \left(\frac{A_k \gamma}{\mu^k}\right)},
	\end{equation}
	where
	\begin{equation}
	\varphi_\omega(x) = x(1-x^{-\omega}) - \omega \log x
	\end{equation}
	is a real function, applied coordinate-wise.
\end{lemma}
\begin{proof}
From \eqref{KL}, we have $F(\gamma^1)-F(\gamma^2)= \sum_{i,j}\left(\gamma^*_{i,j}\log(\gamma^2_{i,j}/\gamma^1_{i,j})+\gamma^1_{i,j}-\gamma^2_{i,j}\right)$. The result can then be deduced from relations \eqref{eq:def_or_proj} and \eqref{scaling}.
\end{proof}
Thus,  the decrease in value of the function $F$ for a over-relaxed projection associated to $\Ccal_k$ is cheap to estimate, since its computational cost is  linear with respect to the dimension of data $\mu^k$. In Figure \ref{phi_omega}, we illustrate the functions  $\varphi_\omega$ for several values of $\omega$.
Notice that for the SK algorithm, which corresponds to $\omega=1$, the function $\varphi_\omega$ is always nonnegative. For other values $1\le\omega<2$, it is nonnegative for arguments close to 1.
\begin{figure}[ht!]
\begin{center}
\includegraphics[width=11cm]{phi_omega.png}
\caption{\label{phi_omega} Plot of functions $\varphi_\omega(x)$ for different values $1\leq \omega\leq 2$ and   $x\in[10^{-2}, 10^1]$.}
\end{center}
\end{figure}
\subsection{Proposed algorithm}

@@

\begin{lemma}[\cite{cuturi13}]
	\label{lemma:trivial_intersection}
	Let us take $\gamma^0$ in $\IR_{+*}^{n_1 n_2}$,
	and denote
	\[
	S = \left\{
	\diag(a) \gamma^0 \diag(b),\quad
	(a,b) \in \IR_{+*}^{n_1 + n_2}
	\right\}
	\]
	the set of matrices that are diagonally similar to $\gamma^0$.
	Then the set $S \cap \Ccal_1 \cap \Ccal_2$ contains exactly one element $\gamma^* = P_{\Ccal_1 \cap \Ccal_2}(\gamma^0)$.
\end{lemma}

@@

\begin{lemma}\label{lemma:F_P_theta}
	Let $1\le \theta < \omega$. Then, for any $\gamma \in \IR_{+*}^{nm}$, one has
	\begin{equation}\label{eq:F_P_theta}
	F(P^\theta_{\Ccal_k}(\gamma)) \le F(P^\omega_{\Ccal_k}(\gamma)).
	\end{equation}
	Moreover, equality occurs if and only if $\gamma \in \Ccal_k$.
\end{lemma}
\begin{proof}
	Thanks to lemma \ref{lemma:lyapunov_decrease}, one knows that
	\[
	F(P^\theta_{\Ccal_k}(\gamma)) - F(P^\omega_{\Ccal_k}(\gamma))
	= \scal{\mu^k}{(\varphi_\omega - \varphi_\theta) \left( \frac{A_k \gamma}{\mu^k} \right) } .
	\]
	Moreover, the function that maps $t \in [1,\infty)$ to $\varphi_t(x)$ is non-increasing:
	$\frac{d}{dt} \varphi_t(x) = \log x (x^{1-t} - 1).$
	For $x\neq 1$, it is even strictly decreasing.
	Thus the inequality (\ref{eq:F_P_theta}) is valid, with equality \emph{iff} $A_k \gamma = \mu$.
\end{proof}

\begin{proposition}
	We argue that a good choice for the functions $\theta_k$ may be constructed as follows. Pick a target parameter $\theta_0 \in [1;2)$, and a small $\delta>0$.
	Define the functions $\Theta^*$ and $\Theta$ as:
	\begin{align}
	\Theta^*(u) &= \sup \left\{\omega \in [1;2]  \mid \varphi_\omega\left(\min u\right) \ge 0 \right\} ,\\
	\label{eq:Theta}
	\Theta(u) &= \min(\max(1,\Theta^*(u)-\delta),\theta_0),
	\end{align}
	where $\min u$ denotes the smallest coordinate of the vector $u$.
	Then the function
	\begin{equation}
	\label{eq:theta_k}
	\theta_k(\gamma) =\Theta\left (\frac{A_k \gamma}{\mu^k}\right)
	\end{equation}
	is continuous, and verifies condition (\ref{eq:cond_theta_k}).
\end{proposition}
This construction has several advantages.
One is that it allows to choose arbitrarily the parameter $\theta_0$ that will be used asymptotically. A heuristic for this choice is proposed in section \ref{section:local}.
But it is also an easy approach to having an adaptive method, as the approximation of $\theta_k^*$ has a negligible cost (it only require to solve a one dimensional problem that depends on the smallest value of $\frac{A_k \gamma}{\mu^k}$, and the corresponding function to invert is decreasing and concave).
We will show later on that this is often an excellent choice in practice.
\begin{proof}
	Since the partial derivative of $\frac{\partial\varphi_\omega(x)}{\partial\omega}$ is nonzero for any $x<1$, the implicit function theorem proves the continuity of $\Theta^*$.
	The function $\Theta^*\left(\frac{A_k \gamma}{\mu^k}\right)$ is such that every term in relation (\ref{eq:kl_diff_scal}) is non-negative.
	Therefore, by lemma \ref{lemma:F_P_theta}, using this parameter minus $\delta$ ensures the strong decrease (\ref{eq:cond_theta_k}) of the Lyapunov function.
	It is no problem to constrain it to $[1,\theta_0]$, as using 1 always verifies the inequality (\ref{eq:cond_theta_k}), and using $\theta_0$ when it is smaller than $\Theta^*\left(\frac{A_k \gamma}{\mu^k}\right)$ does as well.
\end{proof}

@@

\renewcommand{\algorithmiccomment}[1]{\hfill\bgroup(#1)\egroup}
%\begin{algorithm}
%\caption{Over-relaxed SK algorithm}
%\label{SOR}
%\begin{algorithmic}
%\REQUIRE $\mu^1\in \IR^{n_1}$, $\mu^2\in \IR^{n_2}$, $c\in \IR^{n_1\times n_2}_+$
%\STATE Set $a=\mathbf{1}_{n_1}$, $b=\mathbf{1}_{n_2}$, $\gamma^0=e^{-c/\epsilon}$, $\omega\in[1;2[$ and $\eta>0$
%\WHILE {$||a\otimes \gamma^0b-  \mu_1||>\eta$}
%\STATE $\tilde a=\mu_1\oslash (\gamma^0 b)$,  $a_\omega=a^{1-\omega}\otimes \tilde a^\omega$
%\STATE $\tilde b=\mu_2\oslash (^t\gamma^0  a_\omega)$
%\IF{ $\langle \mu^1,\varphi_\omega(a\oslash\tilde a) \rangle +\langle \mu^2,\varphi_\omega(b\oslash\tilde b)\rangle>0$}
%\STATE  $a=a_\omega$, $b=b^{1-\omega}\otimes \tilde b^\omega$\COMMENT{Over-relaxed SK iteration}
%\ELSE
%\STATE $a=\tilde a$, $b=\mu_2\oslash (^t\gamma^0 a)$  \COMMENT{Sinkhorn iteration}
%\ENDIF
%\ENDWHILE
% \RETURN $\gamma=\diag(a)\gamma^0\diag(b)$
%\end{algorithmic}
%\end{algorithm}

@@

\section{Conclusion and perspectives}
Contributions: 
cheap test, preliminary work for convergence results. Huge acceleration in practice.

@@

>>>>

The computation of optimal transport between two data relies on the estimation of an optimal transport matrix, the entries of which represent the quantity of mass transported between  data locations. 
Regularization of optimal transport with strictly convex regularizers \cite{cuturi13, dessein2016}  nevertheless involves a spreading of the mass. Hence, for particular purposes such as color interpolation \cite{Rabin2014} or gradient flow \cite{2016arXiv160705816C}, it is  necessary  to consider very small regularization of the problem.
In this setting,  the regularized transport problem can be ill-conditioned, rendering the Sinkhorn--Knopp algorithm significantly slower. This is the issue  we want to tackle here.
Before going into further details, we now briefly introduce the main notations and concepts used all along this article.

@@

In \cite{cuturi13}, it has been proposed to regularize this problem by adding a strictly convex entropy regularization:
\begin{equation}\label{ROT}
\min_{\gamma\in\Ccal_1\cap \Ccal_2\cap \IR^{n_1 n_2}_{+*}}K^\epsilon(\gamma) := \scal{c}{\gamma} 
+ \epsilon KL(\gamma,\mathbf{1})
,\end{equation}
with $\epsilon>0$, $\mathbf{1}$ is the matrix of size $n_1\times n_2$ full of ones and the Kullback-Leibler divergence is:
\begin{equation}\label{KL}
KL(\gamma,\xi) = \sum_{i,j} \gamma_{i,j} \left( \log \left( \frac{\gamma_{i,j}}{\xi_{i,j}} \right) -1  \right) + \sum_{i,j} \xi_{i,j}.
\end{equation}
It was shown in \cite{benamou15}  that the regularized optimal transport matrix $\gamma^*$, which is the unique minimizer of problem \eqref{ROT},  is the Bregman projection of $\gamma^0 = e^{-c/\epsilon}$ onto $\Ccal_1 \cap \Ccal_2$:
\begin{equation}\label{eq:reg_ot_pb}
\gamma^* = \argmin_{\Ccal_1 \cap \Ccal_2} K^\epsilon(\gamma)= P_{\Ccal_1 \cap \Ccal_2} (e^{-c/\epsilon}),
\end{equation}
where $P_{\Ccal}$ is the  Bregman projection onto $\Ccal$ defined as:
\begin{equation}\label{eq:P_C}
P_{\Ccal}(\xi) := \argmin_{\gamma \in \Ccal} KL(\gamma,\xi).
\end{equation}

@@

In the discrete setting,  projections actually correspond to diagonal scalings. This can be seen by writing the minimality conditions for the projection, given equations (\ref{KL}) and (\ref{eq:P_C}).
\begin{align}\label{scaling}
P_{\Ccal_1}(\gamma) &= \diag(a) \gamma &
a &=  {\mu^1}\oslash{A_1 \gamma} \\
P_{\Ccal_2}(\gamma) &= \gamma \diag(b) &
b &= {\mu^2}\oslash{A_2 \gamma}\nonumber
\end{align}
where $\oslash$ is the pointwise division. 
To compute numerically the solution  one simply has to store $(a^l, b^l)\in\IR^{n_1+n_2}$ and to iterate
\begin{align*}
a^{l+1} &= {\mu^1}\oslash{\gamma^0 b^l} &
b^{l+1} &= {\mu^2}\oslash{^t \gamma^0 a^{l+1}} 
\end{align*}
where the transport matrix can be recovered as :
$\gamma^l = \diag(a^l) \gamma^0 \diag(b^l).$ 
Efficient parallel computations can be considered \cite{cuturi13} and one can even reach real-time computation for large scale problem for certain class of cost matrices $c$ \cite{Solomon2015}. 
For small values of the parameter $\epsilon$, numerical issues can arise and a stabilization of the algorithm is necessary \cite{2016arXiv160705816C}.
The convergence of the process can nevertheless be very slow  in the setting $\epsilon\to 0$.
\subsection{Related works }
The introduction of relaxation variables through heavy ball approaches \cite{POLYAK19641} has recently gained in popularity  to speed up the convergence of algorithms optimizing convex \cite{2014arXiv1412.7457G} or non convex \cite{Zavriev1993,2016arXiv160609070O} problems. Such schemes have also been empirically considered to accelerate the SK algorithm  in \cite{peyre2016quantum,2017arXiv170801955S}. The convergence of these algorithms is nevertheless not studied yet in the context of regularized optimal transport.

@@

As illustrated in Figure \ref{alternate_projections} (a-b), the SK algorithm, that  performs alternate Bregman projections onto the affine sets $\Ccal_1$ and $\Ccal_2$, can be very slow when $\epsilon\to 0$, as the problem becomes ill-conditioned. The idea developped in this paper is to perform over-relaxed projections in order to accelerate the process, as displayed in Figure \ref{alternate_projections} (c).

@@

Let us define the $\omega$-over-relaxed projection operator by:
\begin{equation}\label{eq:def_or_proj}
\log P^\omega_{\Ccal_k}(\gamma) = (1-\omega) \log \gamma + \omega \log P_{\Ccal_k}(\gamma),
\end{equation}
where the logarithm is taken coordinate-wise.
Note that $P_{\Ccal_k}^0$ is the identity, $P_{\Ccal_k}^1 = P_{\Ccal_k}$ is the standard Bregman projection, and $P_{\Ccal_k}^2$ is an involution (recall that $\Ccal_k$ is an affine subspace).

@@

We argue that it makes sense to consider this Lyapunov function.
Obviously, it is closely related to the notion of Bregman projection that is used throughout the algorithm.
But it is also easy to calculate its decrease, as shown in lemma \ref{lemma:lyapunov_decrease}.
Moreover, as we will see later, it allows the use of a wide range of over-relaxation parameters in practice.

\begin{lemma} \label{lemma:KL_compact}
	For any $M \in \IR_+^*$, the sublevel set $K := \left\{ \gamma \mid F(\gamma) \le M \right\}$ is compact.
\end{lemma}
\begin{proof}
	First note that K is closed in $\IR_{+*}^{n_1 n_2}$, by continuity of $F$.
	For $\gamma \in K$, one has:
	\begin{align*}
	F(\gamma) &= \scal{\gamma^*}{\log \gamma^* - 1} + \scal{\gamma}{1} - \scal{\gamma^*}{\log \gamma}\\
	\sum_{i,j} \gamma_{i,j} - \gamma^*_{i,j} \log \gamma_{i,j} &\le M - \scal{\gamma^*}{\log \gamma^* - 1} .
	\end{align*}
	By concavity, the logarithm admits a simple upper bound: $\log \gamma_{i,j} \le \frac{\gamma_{i,j}}{e}$. Moreover, since $\gamma^*$ is a solution of the problem, each of its coordinates is at most 1.
	\begin{align*}
	\sum_{i,j} \gamma_{i,j} &\le \frac{M - \scal{\gamma^*}{\log \gamma^*-1}}{1-\frac{1}{e}} =: M'.
	\end{align*}
	Therefore each coordinate $\gamma_{i,j}$ is bounded above by $M'$.
	Let us show that the coordinates are bounded below by a strictly positive number; from which can be deduced that $K$ is bounded and closed in $\IR^{n_1 n_2}$, and therefore compact.
	\begin{align*}
	F(\gamma)
	&\ge \scal{\gamma^*}{\log \gamma^*} - \sum_{i,j} \gamma^*_{i,j} \log(\gamma_{i,j}) - 1\\
	&\ge \scal{\gamma^*}{\log \gamma^*} - \gamma^*_{i_0,j_0} \log(\gamma_{i_0,j_0}) - 1 - \log M',
	\end{align*}
	where the last equation derives from $\log(\gamma_{i,j}) \le \log M'$, and $\sum_{(i,j)\neq(i_0,j_0)} \gamma^*_{i,j} \le 1$.
	Finally:
	\[
	\log(\gamma_{i_0,j_0}) \ge \frac{\scal{\gamma^*}{\log \gamma^*} - M - \log M' - 1}{\gamma^*_{i_0,j_0}},
	\]
	and $K$ is thus compact.
\end{proof}

Note that the difference $F(\gamma) - F(P^\omega_{\Ccal_k}(\gamma))$ may be calculated without knowing $\gamma^*$, as shown by the following lemma.
\begin{lemma}\label{lemma:lyapunov_decrease}
	Take $\gamma$ in $\IR^{mn}_{+*}$. The decrease of the Lyapunov function when applying an over-relaxed projection can be calculated with the following formula:
	\begin{equation} \label{eq:kl_diff_scal}
	F(\gamma) - F(P^\omega_{\Ccal_k}(\gamma)) = 
	\scal{\mu^k}{\varphi_\omega \left(\frac{A_k \gamma}{\mu^k}\right)},
	\end{equation}
	where
	\begin{equation}
	\varphi_\omega(x) = x(1-x^{-\omega}) - \omega \log x
	\end{equation}
	is a real function, applied coordinate-wise.
\end{lemma}
\begin{proof}
From \eqref{KL}, we have $F(\gamma^1)-F(\gamma^2)= \sum_{i,j}\left(\gamma^*_{i,j}\log(\gamma^2_{i,j}/\gamma^1_{i,j})+\gamma^1_{i,j}-\gamma^2_{i,j}\right)$. The result can then be deduced from relations \eqref{eq:def_or_proj} and \eqref{scaling}.
\end{proof}
Thus, the decrease of the Lyapunov function $F$ for an over-relaxed projection associated to $\Ccal_k$ is cheap to estimate, since its computational cost is  linear with respect to the dimension of data $\mu^k$. In Figure \ref{phi_omega}, we illustrate the functions  $\varphi_\omega$ for several values of $\omega$.
Notice that for the SK algorithm, which corresponds to $\omega=1$, the function $\varphi_\omega$ is always non-negative. For other values $1\le\omega<2$, it is non-negative for arguments close to 1.
\begin{figure}[ht!]
\begin{center}
\includegraphics[width=11cm]{phi_omega.png}
\caption{\label{phi_omega} Plot of functions $\varphi_\omega(x)$ for different values $1\leq \omega\leq 2$ and   $x\in[10^{-2}, 10^1]$.}
\end{center}
\end{figure}
\subsection{Proposed algorithm}

@@

\begin{lemma}[]
	\label{lemma:trivial_intersection}
	Let us take $\gamma^0$ in $\IR_{+*}^{n_1 n_2}$,
	and denote
	\[
	S = \left\{
	\diag(a) \gamma^0 \diag(b),\quad
	(a,b) \in \IR_{+*}^{n_1 + n_2}
	\right\}
	\]
	the set of matrices that are diagonally similar to $\gamma^0$.
	Then the set $S \cap \Ccal_1 \cap \Ccal_2$ contains exactly one element $\gamma^* = P_{\Ccal_1 \cap \Ccal_2}(\gamma^0)$.
\end{lemma}
We refer to \cite{cuturi13} for the proof of this well-known result, which is used in the proof of the theorem.

@@

\begin{lemma}\label{lemma:F_P_theta}
	Let $1\le \theta < \omega$. Then, for any $\gamma \in \IR_{+*}^{nm}$, one has
	\begin{equation}\label{eq:F_P_theta}
	F(P^\theta_{\Ccal_k}(\gamma)) \le F(P^\omega_{\Ccal_k}(\gamma)).
	\end{equation}
	Moreover, equality occurs if and only if $\gamma \in \Ccal_k$.
\end{lemma}
\begin{proof}
	Thanks to lemma \ref{lemma:lyapunov_decrease}, one knows that
	\[
	F(P^\theta_{\Ccal_k}(\gamma)) - F(P^\omega_{\Ccal_k}(\gamma))
	= \scal{\mu^k}{(\varphi_\omega - \varphi_\theta) \left( \frac{A_k \gamma}{\mu^k} \right) } .
	\]
	Moreover, the function that maps $t \in [1,\infty)$ to $\varphi_t(x)$ is non-increasing:
	$\frac{d}{dt} \varphi_t(x) = \log x (x^{1-t} - 1).$
	For $x\neq 1$, it is even strictly decreasing.
	Thus the inequality (\ref{eq:F_P_theta}) is valid, with equality \emph{iff} $A_k \gamma = \mu^k$.
\end{proof}

\begin{proposition}
	We argue that a good choice for the functions $\theta_k$ may be constructed as follows. Pick a target parameter $\theta_0 \in [1;2)$, and a small $\delta>0$.
	Define the functions $\Theta^*$ and $\Theta$ as:
	\begin{align}
	\label{eq:Theta_opt}
	\Theta^*(u) &= \sup \left\{\omega \in [1;2]  \mid \varphi_\omega\left(\min u\right) \ge 0 \right\} ,\\
	\label{eq:Theta}
	\Theta(u) &= \min(\max(1,\Theta^*(u)-\delta),\theta_0),
	\end{align}
	where $\min u$ denotes the smallest coordinate of the vector $u$.
	Then the function
	\begin{equation}
	\label{eq:theta_k}
	\theta_k(\gamma) =\Theta\left (\frac{A_k \gamma}{\mu^k}\right)
	\end{equation}
	is continuous, and verifies condition (\ref{eq:cond_theta_k}).
\end{proposition}
This construction has several advantages.
One is that it allows to choose arbitrarily the parameter $\theta_0$ that will be used asymptotically. A heuristic for this choice is proposed in section \ref{section:local}.
But it is also an easy approach to having an adaptive method, as the approximation of $\theta_k^*$ has a negligible cost (it only require to solve a one dimensional problem that depends on the smallest value of $\frac{A_k \gamma}{\mu^k}$, and the corresponding function to invert is decreasing and concave).
We will show later on that this is often an excellent choice in practice.
\begin{proof}
	Since the partial derivative of $\frac{\partial\varphi_\omega(x)}{\partial\omega}$ is nonzero for any $x<1$, the implicit function theorem proves the continuity of $\Theta^*$.
	The function $\Theta^*\left(\frac{A_k \gamma}{\mu^k}\right)$ is such that every term in relation (\ref{eq:kl_diff_scal}) is non-negative.
	Therefore, by lemma \ref{lemma:F_P_theta}, using this parameter minus $\delta$ ensures the strong decrease (\ref{eq:cond_theta_k}) of the Lyapunov function.
	It is no problem to constrain it to $[1,\theta_0]$, as using 1 always verifies the inequality (\ref{eq:cond_theta_k}), and using $\theta_0$ when it is smaller than $\Theta^*\left(\frac{A_k \gamma}{\mu^k}\right)$ does as well.
\end{proof}

@@

\renewcommand{\algorithmiccomment}[1]{\hfill\bgroup(#1)\egroup}
%\begin{algorithm}
%\caption{Over-relaxed SK algorithm}
%\label{SOR}
%\begin{algorithmic}
%\REQUIRE $\mu^1\in \IR^{n_1}$, $\mu^2\in \IR^{n_2}$, $c\in \IR^{n_1\times n_2}_+$
%\STATE Set $a=\mathbf{1}_{n_1}$, $b=\mathbf{1}_{n_2}$, $\gamma^0=e^{-c/\epsilon}$, $\omega\in[1;2[$ and $\eta>0$
%\WHILE {$||a\otimes \gamma^0b-  \mu_1||>\eta$}
%\STATE $\tilde a=\mu_1\oslash (\gamma^0 b)$,  $a_\omega=a^{1-\omega}\otimes \tilde a^\omega$
%\STATE $\tilde b=\mu_2\oslash (^t\gamma^0  a_\omega)$
%\IF{ $\langle \mu^1,\varphi_\omega(a\oslash\tilde a) \rangle +\langle \mu^2,\varphi_\omega(b\oslash\tilde b)\rangle>0$}
%\STATE  $a=a_\omega$, $b=b^{1-\omega}\otimes \tilde b^\omega$\COMMENT{Over-relaxed SK iteration}
%\ELSE
%\STATE $a=\tilde a$, $b=\mu_2\oslash (^t\gamma^0 a)$  \COMMENT{Sinkhorn iteration}
%\ENDIF
%\ENDWHILE
% \RETURN $\gamma=\diag(a)\gamma^0\diag(b)$
%\end{algorithmic}
%\end{algorithm}
Add algorithm to solve \eqref{eq:Theta_opt}?

@@

\section{Conclusion and perspectives}

We have presented a generalization of the Sinkhorn--Knopp algorithm based on over-relaxed projections, with a criterion for global convergence based on a Lyapunov function.
The proposed Lyapunov function allows for an adaptive choice of over-relaxation parameter at a low computational price.
In practice, this algorithm converges way faster than Sinkhorn--Knopp; moreover, the acceleration rate is especially impressive for small regularization $\epsilon$.

Contributions: 
cheap test, preliminary work for convergence results. Huge acceleration in practice.

@@

Links with monotone Fista? (that choose Fista if energy decreased and forward backward otherwise...)
